Parsing Postfix log files

### 2007-12-23 doc/logparser.tex
Improved the discussion of why rules and actions are separated, by incorporating the appropriate paragraph from the rules section; the rules and algorithm sections now refer to the parser design section. Extended the conclusion somewhat.

@@

@@

The complexity and variation in Postfix's logs requires similar flexibility
in the parser, which is split into two parts: the parsing algorithm and the
parsing rules which are applied to the lines.  Decoupling the parsing rules
from the associated actions allows new rules to be written and tested
without requiring modifications to the algorithm source code (significantly
lowering the barrier to entry for new or casual users who need to parse new
log lines), and greatly simplifies both algorithm and rules.  Decoupling
also creates a clear separation of functionality: rules handle low level
details of identifying log lines and extracting data from a line, whereas
the algorithm handles the higher level details of following the path a mail
takes through Postfix, assembling the required data before storing it,
dealing with complications arising, etc.

@@

Parsing Postfix logs appears at first sight to be an almost trivial task,
especially given previous experience in parsing logs, but it turns out to
be a much more taxing project than initially expected.  The variety and
breadth of log lines produced by Postfix is quite surprising, as a quick
survey of sample log files gives the impression that the number of distinct
log lines is quite small; this impression is due to the uneven distribution
evidenced by logs produced in normal operation (see graph~\refwithpage{rule
hits graph} for a vivid illustration of this).

Enabling users to easily extend the parser to deal with new log lines is a
design imperative, and despite the resulting increase in complexity the job
is quite tractable, though it does raise interesting efficiency and
optimisation questions (answered in section~\refwithpage{rule efficiency}).
For the parser to be useful to the average user, parsing new log lines
produced by Postfix must be as simple a task as configuring Postfix to
produce those log lines was in the first place, otherwise users won't make
to effort to utilise the parser.  Providing a tool to ease the generation
of \regexes{} from unparsed log lines should help

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications previously
explained only begin to become apparent once the vast majority of log lines
have been parsed successfully.  Dealing with numerous, infrequently
occurring log lines is a simple, albeit tiresome task, whereas dealing with
information which hasn't been logged, e.g.\ when delivery attempts are
aborted.

>>>>

Decoupling the parsing rules from the associated actions allows new rules
to be written and tested without requiring modifications to the algorithm
source code (significantly lowering the barrier to entry for new or casual
users who need to parse new log lines), and greatly simplifies both
algorithm and rules.  Decoupling also creates a clear separation of
functionality: rules handle low level details of identifying log lines and
extracting data from a line, whereas the algorithm handles the higher level
details of following the path a mail takes through Postfix, assembling the
required data before storing it, dealing with complications arising, etc.

@@

Please refer to section~\refwithpage{parser design} for a discussion of why
the rules and actions have been separated in the parser's design.

@@

@@

Parsing Postfix logs appears at first sight to be an almost trivial task,
especially given previous experience in parsing logs, but it turns out to
be a much more taxing project than initially expected.  The variety and
breadth of log lines produced by Postfix is quite surprising, as a quick
survey of sample log files gives the impression that the number of distinct
log lines is quite small; this impression is due to the uneven distribution
exhibited by logs produced in normal operation (see graph~\refwithpage{rule
hits graph} for a vivid illustration of this).

Enabling users to easily extend the parser to deal with new log lines is a
design imperative, and despite the resulting increase in complexity the job
is quite tractable, though it does raise interesting efficiency and
optimisation questions (answered in section~\refwithpage{rule efficiency}).
For the parser to be useful to the average user, parsing new log lines
produced by Postfix must be as simple a task as configuring Postfix to
produce those log lines was in the first place, otherwise users won't make
to effort to utilise the parser.  Providing a tool to ease the generation
of \regexes{} from unparsed log lines should greatly help users
add rules to parse unparsed lines.

The real difficulties arise once the parser is successfully dealing with
90\% of the log lines, as the irregularities and complications previously
explained only begin to become apparent once the vast majority of log lines
have been parsed successfully.  Dealing with numerous, infrequently
occurring log lines is a simple task, albeit tiresome, whereas dealing with
mails where information appears to be missing is much more grueling.
Trawling through the logs, looking for something out of the ordinary,
possibly hundreds or even thousands of lines away from the last mention of
the queueid in question, is extremely time consuming and error prone.  In
many cases the task is not to spot the line which is unusual, but to spot
that a line normally present is missing.  In all cases the evidence must be
used to construct a hypothesis to explain the irregularities, and that
hypothesis must then be tested in the parser; if successful the parser must
be modified to deal with the irregularities, without adversely affecting
the existing functionality.  The complications described in
section~\refwithpage{additional complications} were solved in the order
they are described in, and that order closely resembles the frequency with
which they occur; the most frequently occurring complications dominate the
warning messages produced, and so naturally they are the first
complications to be dealt with.
